Safety Cases for Software Product Lines

### 2017-10-07 doc/SafetyCases/SafetyCasesPL.tex
Generative, annotative and delta-oriented pseudo-code

\begin{enumerate}
\item Truck with liquid fuel
\item Bus with liquid fuel
\item Truck with gas
\item Bus with gas
\end{enumerate}

section{Software Product Lines (SPLs)}

@@

\begin{enumerate}
\item FuelLevel
\item FuelWarning (not present in trucks with gas)
\item Truck
\item Bus
\item LiquidFuel
\item Gas
\end{enumerate}

@@

\item[Presence Condition]
A Presence Condition (PC) is a boolean formula over the set of features that does not conflict with the Feature Model.
\end{description}

For example, an FLD product for buses, with liquid fuel, with fuel level display and low fuel warning can be represented by the following PC:

Bus $\wedge$ Liquid $\wedge$ Estimation $\wedge$ Warning

@@

\begin{figure}
  \centering
    \includegraphics[width=0.75\textwidth]{FeatureOriented}
  \caption{Feature-Oriented implementation of the object store Product Line}
  \label{fig:FeatureOriented}
\end{figure}

Figure \ref{fig:FeatureOriented} shows a possible Feature-Oriented implementation of our object store Product Line. \emph{SingleStore} and \emph{MultiStore} are two separate implementations of a their corresponding features, and \emph{AccessControl} is implemented as a cross-cutting feature that refines any of the 2 \emph{Store} implementations.

\begin{figure}
  \centering
    \includegraphics[width=0.75\textwidth]{FeatureOrientedGenerated}
  \caption{The product generated out of the \emph{MultiStore} and \emph{AccessControl} features}
  \label{fig:FeatureOrientedGenerated}
\end{figure}

A product can then be generated out of the individual features given a valid Presence Condition. For example, if we are to generate a product with the \emph{MultiStore} and \emph{AccessControl} features, the generator would automatically generate the code in figure \ref{fig:FeatureOrientedGenerated}.

@@

If we are to rewrite our object store Product Line using CPP macros, it would look like listing 
ref{lst:Annotative}. A single feature (e.g. \emph{ACCESS\_CONTROL} is now scattered across the code base, and in some cases replicated (e.g. lines 13-15 and 24-26).

\begin{lstlisting}[language=Java, caption=Annotative example, label={lst:Annotative}]
class Store {
	#if defined (ACCESS_CONTROL)
	private boolean sealed = false;
	#endif

	#if defined (SINGLE_STORE)
	private Object value;
	Object read() {
		#if defined (ACCESS_CONTROL)
 		if (!sealed) {
		#endif
			return value;
		#if defined (ACCESS_CONTROL)
		} else { throw new RuntimeException("Access denied!"); }
		#endif 
	}
	#elif defined (MULTI_STORE)
	private LinkedList values = new LinkedList();
	Object read() { 
		#if defined (ACCESS_CONTROL)
 		if (!sealed) {
		#endif
			return values.getFirst(); 
		#if defined (ACCESS_CONTROL)
		} else { throw new RuntimeException("Access denied!"); }
		#endif
	}
	Object readAll() { return values.toArray(); }
	#endif

	void set(Object nvalue) {
		#if defined (ACCESS_CONTROL)
 		if (!sealed) {
		#endif
			#if defined (SINGLE_STORE)
			value = nvalue;
			#elif defined (MULTI_STORE)
			values.addFirst(nvalue);
			#endif
		#if defined (ACCESS_CONTROL)
		} else { throw new RuntimeException("Access denied!"); }
		#endif
	}
}
\end{lstlisting}

@@

Listing \ref{lst:Annotative} shows a Delta-Oriented implementation of our object store Product Line. The base product here is a simple single-object store. On top of that, delta for a multi-object store and an access controlled store are defined as deltas.

\begin{lstlisting}[caption=Annotative example, label={lst:Annotative}]
core SingleStore {
	class Store {
		private Object value;
		Object read() { return value; }
		void set(Object nvalue) { value = nvalue; }
	} 
}

delta DMultiStore when MultiStore {
	modifies class Store {
		remove Object value;
		add private LinkedList values = new LinkedList();

		remove read();
		add Object read() { return values.getFirst(); }

		add Object[] readAll() { return values.toArray(); }

		remove void set(Object);
		add void set(Object nvalue) { values.addFirst(nvalue); }
	}
}

delta DAccessControl when AccessControl {
	modifies class Store {
		add private boolean sealed = false;

		remove read();
		add Object read() {
			if (!sealed) {
				return Store.read();
			} else { throw new RuntimeException("Access denied!"); }
		}
	}
}
\end{lstlisting}

>>>>

\begin{enumerate}
\item Truck with liquid fuel
\item Bus with liquid fuel
\item Truck with gas (doesn't include the fuel level warning feature)
\item Bus with gas
\end{enumerate}

For vehicles with gas, the fuel level is reported directly to the driver. However, liquid fuel level in the tank fluctuates due to vehicle movement. To smooth the liquid gas level reported to the driver, the level read from the tank sensor is passed into a filter. Two types of filters are used:

\begin{enumerate}
\item a Kalman filter is used in trucks
\item a Low-Pass filter is used in buses
\end{enumerate}

\section{Software Product Lines (SPLs)}

@@

\begin{enumerate}
\item FuelLevel
\item FuelWarning (not present in trucks with gas)
\item Truck
\item Bus
\item Liquid
\item Gas
\end{enumerate}

@@

\item[Presence Condition]
A Presence Condition (PC) is a boolean formula over the set of features that represents the set of products in which an artifact exists.
\end{description}

For example, the presence condition for the Low Fuel Level Warning feature is:

$\neg ($ Truck $\wedge$ Gas $)$

@@

\begin{lstlisting}[language=Java, caption=Generative example, label={lst:Generative}]
// Feature module FLDGas
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(level); 
	}
}

// Feature module FLDLiquidTruck
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(filterKalman(level)); 
	}
}

// Feature module FLDLiquidBus
class FLD {
	void reportFuelLevel(float level) { 
		dashboard.display(filterLowPass(level)); 
	}
}

// Feature module LowGasWarning
refines class FLD {
	const float THRESHOLD;

	void reportFuelLevel(float level) { 
		if (level < THRESHOLD) { dashboard.triggerWarning()); }
		Super.reportFuelLevel(level);
	}
} 
\end{lstlisting}

Listing \ref{lst:Generative} shows a Feature-Oriented high-level implementation of FLDPL. \emph{FLDGas}, \emph{FLDLiquidTruck} and \emph{FLDLiquidBus} are three separate implementations of the Fuel Level Display module. \emph{LowGasWarning} is implemented as a cross-cutting feature that refines any of the 3 \emph{FLD} implementations.

\begin{lstlisting}[language=Java, caption=Generated Product, label={lst:Generated}]
class FLD {
	const float THRESHOLD;

	void reportFuelLevel(float level) { 
		if (level < THRESHOLD) { dashboard.triggerWarning()); }
		dashboard.display(filterKalman(level);
	}
} 
\end{lstlisting}

Given the feature expression (Liquid $\wedge$ Truck $\wedge$ FuelWarning), the product in listing \ref{lst:Generated} is generated.

@@

If we are to rewrite FLDPL using CPP macros, it would look like listing 
ref{lst:Annotative}. A single feature (e.g. \emph{LOW\_GAS\_WARNING} is now scattered across the code base, and in some cases replicated.

\begin{lstlisting}[language=Java, caption=Annotative example, label={lst:Annotative}]
class FLD {
	#if defined (LOW_GAS_WARNING)
	const float THRESHOLD;
	#endif

	void reportFuelLevel(float level) {
		#if defined (LOW_GAS_WARNING)
		if (level < THRESHOLD) { dashboard.triggerWarning(); }
		#endif

		dashboard.display(
			#if defined (FLD_GAS)
			level
			#endif

			#if defined (FLD_LIQUID_TRUCK)
			filterKalman(level);
			#endif

			#if defined (FLD_LIQUID_BUS)
			filterLowPass(level);
			#endif
		);
	}
}
\end{lstlisting}

@@

Listing \ref{lst:Delta} shows a Delta-Oriented implementation of FLDPL. The base product here is the Gas FLD. On top of that, deltas for other variants are implemented.

\begin{lstlisting}[caption=Delta-Oriented example, label={lst:Delta}]
core FLDGas {
	class FLD {
		void reportFuelLevel(float level) { 
			dashboard.display(level); 
		}
	} 
}

delta DLiquidTruck when LiquidTruck {
	modifies class FLD {
		remove reportFuelLevel;
		add void reportFuelLevel(float level) { 
			dashboard.display(filterKalman(level)); 
		}
}

delta DLiquidBus when LiquidBus {
	modifies class FLD {
		remove reportFuelLevel;
		add void reportFuelLevel(float level) { 
			dashboard.display(filterLowPass(level)); 
		}
}

delta DLowGasWarning when LowGasWarning {
	modifies class FLD {
		add const float THRESHOLD;

		remove reportFuelLevel;
		add void reportFuelLevel(float level) {
			if (level < THRESHOLD) { dashboard.triggerWarning(); }
			FLD.reportFuelLevel(level);
		}
	}
}
\end{lstlisting}
